---
title: "Offshore Marine Coral Analysis"
subtitle: "A Healthy Waters Partnership Analysis"
description: "This script analyses, scores and presents coral data in the offshore marine environment in the Dry Tropics region."
author: "Adam Shand"
format: html
params:
  project_crs: "EPSG:7844"
  target_fyear: 2024
---

# Introduction

This is a simple script to produce a map of sampling locations for the Coral indicator category in the Offshore Marine Zone for the Dry Tropics region.

# Script Set Up

This script requires multiple core spatial packages, each of which is loaded below. Key variables and paths are also created.

```{r}
#| label: load packages

#use pacman function to load and install (if required) all other packages
pacman::p_load(tidyverse, glue, here, sf, tmap, janitor, dataaimsr)

#load in the custom function used to create the read and write folders for the script
source(here("functions/script_setup.R"))

#run the function to create the folders and paths
script_setup()

#set project crs
proj_crs <- params$project_crs

#read in the custom function to clean column names into our specific style
source(here("functions/name_cleaning.R"))

```

# Load Data

Next we load in the data, refer to the README in the GitHub repo if data is missing.

```{r}
#| label: load in wetland data

#please note if the data is not in the basin builder folder, try running the basin builder script
dry_tropics <- n3_region |> 
  name_cleaning() |> 
  st_transform(proj_crs) |> 
  filter(Environment == "Marine", Region == "Dry Tropics")

#read in reef data and update crs
reefs <- get(data("gbr_feat", package = "gisaimsr")) |> 
  name_cleaning() |> 
  filter(FeatName == "Reef") |> 
  st_transform(proj_crs)

#create a townsville point coord
tsv <- st_as_sf(data.frame(place = "Townsville", x = "-19.2590", y = "146.8169"), coords = c("y", "x"), crs = proj_crs)

```

# Maipulate Data

The names in the reef dataset are too excessive and can visually clutter our map, in this code chunk we clean up the reef names.

```{r}
#| label: clean up data

#crop reefs to only in Dry Tropics region and take only polygons
temp_reefs <- st_intersection(reefs, dry_tropics) |> st_collection_extract("POLYGON")

#create list of reefs we care about and their x and y label modifications. Order is Name = Xmod, Ymod)
reef_list <- list("Chicken Reef" = c(1, 1.2),
                  "Davies Reef" = c(0, 0.8),
                  "Dip Reef" = c(0, 0.8),
                  "Helix Reef" = c(0, -0.5),
                  "John Brewer Reef" = c(-1, 0.8),
                  "Kelso Reef" = c(0, 0.8),
                  "Knife Reef" = c(0, 0.8), 
                  "Myrmidon Reef" = c(0, 0.8),
                  "Rib Reef" = c(-0.5, 0.7))

#filter by name to only get the offshore reefs we want, group, summarise and mutate reef data
dt_reefs <- temp_reefs |> filter(GbrName %in% names(reef_list)) |> select(GbrName) |> 
  mutate(Zone = "offshore", GbrName = as.character(GbrName)) |> 
  rename(Reef = GbrName) |> mutate(Xmod = NA, Ymod = NA)

#loop over the list and add each item
for (i in 1:length(reef_list)){
  
  dt_reefs <- dt_reefs |> mutate(Xmod = case_when(Reef == names(reef_list)[i] ~ reef_list[[i]][1], T ~ Xmod),
                                 Ymod = case_when(Reef == names(reef_list)[i] ~ reef_list[[i]][2], T ~ Ymod))
}

#get only the offshore zone
offshore <- dry_tropics |> filter(BasinOrZone == "Offshore")

```

# Create Map

And finally we can produce the required maps to be used in our technical report style.

```{r}
#| label: create map

#load in custom function to create the inset map
source(here("functions/maps_inset_layer.R"))

#create a background of the burdekin region for the inset map
dt_background <- dry_tropics |> 
  group_by(BasinOrZone) |> 
  summarise(geom = st_union(geom))

#create the background layer using the broader dry tropics backdrop
maps_inset_layer(supplied_sf_1 = offshore,  
                 use_bbox_1 = FALSE,
                 background = dt_background, aspect = 1)

#create a map of the area
map <- tm_shape(qld) +
  tm_polygons(fill = "grey80", 
              col = "black") +
  tm_shape(dry_tropics, is.main = T) +
  tm_polygons(fill = NULL,
              col = "black") +
  tm_shape(tsv) +
  tm_symbols(size = 0.5, 
             fill = "white", 
             col = "black", 
             lwd = 2, 
             shape = 23) +
  tm_text("place", 
          options = opt_tm_text(shadow = T), 
          xmod = -2.5, 
          ymod = 0.1) +
  tm_shape(reefs) +
  tm_borders(col = "grey60",
             col_alpha = 0.4,
             fill = "grey60",
             fill_alpha = 0.2) +
  tm_shape(dt_reefs) +
  tm_polygons(fill = "Reef", 
              fill.scale = tm_scale_categorical(values = "brewer.set3"), 
              fill.legend = tm_legend(show = F),
              col = "black") +
  tm_shape(dt_reefs) +
  tm_text("Reef",
          options = opt_tm_text(shadow = T), 
          xmod = "Xmod", 
          ymod = "Ymod", 
          size = 0.8) +
  tm_layout(legend.bg.color = "white", 
            legend.position = c("left", "bottom"),
            asp = 1)

#save the map as a png
tmap_save(map, filename = glue("{output_path}/coral_offshore-marine.png"), 
            insets_tm = inset_map, insets_vp = inset_viewport)

```

Script complete :) 