---
title: "Offshore Marine Coral Analysis"
subtitle: "A Healthy Waters Partnership Analysis"
description: "This script analyses, scores and presents coral data in the offshore marine environment in the Dry Tropics region."
author: "Adam Shand"
format: html
params:
---

# Introduction

This is a simple script to produce a map of sampling locations for the Coral indicator category in the Offshore Marine Zone for the Dry Tropics region. All scores are precalculated by the data provider.

# Script Set Up

This script requires multiple core spatial packages, each of which is loaded below. Key variables and paths are also created.

```{r}
#| label: load packages

#use pacman function to load and install (if required) all other packages
pacman::p_load(dplyr, glue, here, sf, tmap)

#update/load custom package
#pak::pak("add-am/RcTools")
library(RcTools)

#define the crs targeted
proj_crs <- "EPSG:7855"

#define data paths and output paths
data_path <- glue("{here()}/data")
output_path <- glue("{here()}/outputs")

```

# Load Data

Next we load in the data, this consists of the northern three region outline, and a reefs dataset.

## Northern Three Dataset

```{r}
#| label: load in wetland data

#create a path to where the data should be
n3_reg_path <- glue("{data_path}/n3_region.gpkg")

#either load it in, or create it and save it
if (file.exists(n3_reg_path)){

  n3_region <- st_read(n3_reg_path)

} else {

  n3_region <- build_n3_region()
  st_write(n3_region, n3_reg_path)

}

#restrict to just the dry tropics region
dry_tropics <- n3_region |> 
  filter(Region == "Dry Tropics")

```

## Reef Outlines

```{r}

#read in reef data and update crs
reefs <- get(data("gbr_feat", package = "gisaimsr")) |> 
  name_cleaning() |> 
  filter(FeatName == "Reef") |> 
  st_transform(proj_crs)

```

# Maipulate Data

The names in the reef dataset are too excessive and can visually clutter our map, in this code chunk we clean up the reef names.

```{r}
#| label: clean up data

#crop reefs to only in Dry Tropics region and take only polygons
temp_reefs <- st_intersection(reefs, dry_tropics) |> st_collection_extract("POLYGON")

#create list of reefs we care about and their x and y label modifications. Order is Name = Xmod, Ymod)
reef_list <- list(
  "Chicken Reef" = c(1, 1.2),
  "Davies Reef" = c(0, 0.8),
  "Dip Reef" = c(0, 0.8),
  "Helix Reef" = c(0, -0.5),
  "John Brewer Reef" = c(-1, 0.8),
  "Kelso Reef" = c(0, 0.8),
  "Knife Reef" = c(0, 0.8), 
  "Myrmidon Reef" = c(0, 0.8),
  "Rib Reef" = c(-0.5, 0.7))

#filter by name to only get the offshore reefs we want, group, summarise and mutate reef data
dt_reefs <- temp_reefs |> filter(GbrName %in% names(reef_list)) |> select(GbrName) |> 
  mutate(Zone = "offshore", GbrName = as.character(GbrName)) |> 
  rename(Reef = GbrName) |> mutate(Xmod = NA, Ymod = NA)

#loop over the list and add each item
for (i in 1:length(reef_list)){
  
  dt_reefs <- dt_reefs |> mutate(
    Xmod = case_when(
      Reef == names(reef_list)[i] ~ reef_list[[i]][1], 
      TRUE ~ Xmod),
    Ymod = case_when(
      Reef == names(reef_list)[i] ~ reef_list[[i]][2], 
      TRUE ~ Ymod))
}

```

# Create Map

And finally we can produce the required maps to be used in our technical report style.

```{r}
#| label: create map

#create the water layer
water_layer <- maps_water_layer(c("Ross", "Black"), WaterLines = TRUE, WaterAreas = TRUE, WaterLakes = TRUE, StreamOrder = 2)

#create a background of the dry tropics region for the inset map
dt_background <- dry_tropics |> 
  filter(Environment == "Marine")

#create a background for the main map
dt_land <- dry_tropics |>
  filter(Environment != "Marine") |> 
  group_by(BasinOrZone) |> 
  summarise(geom = st_union(geom))

#create the inset objects
inset_objects <- maps_inset_layer(dt_reefs, dt_background, 1.1)

#create a map of the area
coral_map <- tm_shape(qld) +
  tm_polygons(fill = "grey80", col = "black") +
  tm_shape(dt_background, is.main = TRUE) +
  tm_polygons(fill = NULL, col = "black") +
  tm_shape(dt_land) +
  tm_polygons(fill = "grey90", col = "black") +
  water_layer +
  tm_shape(reefs) +
  tm_borders(
    col = "grey60",
    col_alpha = 0.4,
    fill = "grey60",
    fill_alpha = 0.2) +
  tm_shape(dt_reefs) +
  tm_polygons(
    fill = "Reef",
    fill.scale = tm_scale_categorical(values = "brewer.set3"), 
    fill.legend = tm_legend(show = FALSE),
    col = "black") +
  tm_shape(dt_reefs) +
  tm_text(
    "Reef",
    options = opt_tm_text(shadow = TRUE), 
    xmod = "Xmod", 
    ymod = "Ymod", 
    size = 0.8) +
  tm_layout(
    legend.bg.color = "white", 
    legend.position = c("left", "bottom"),
    asp = 1.1)

#save the map as a png
tmap_save(
  coral_map, 
  filename = glue("{output_path}/coral_offshore-marine.png"),
  insets_tm = inset_objects[1], 
  insets_vp = inset_objects[2])

```

```{r}


```