---
title: "Inshore Coral Script 2 - Data Visualisation"
subtitle: "A Healthy Waters Partnership Analysis"
description: "Script 2 in a series of scripts designed to analyse, score, and present inshore marine coral data in the Dry Tropics region."
author: "Adam Shand"
format: html
params:
  project_crs: "EPSG:7844"
  target_fyear: 2024
---

# Introduction

This is a simple script to produce a map of sampling locations for the Coral indicator category in the Inshore Marine Zone for the Dry Tropics region.

# Script Set Up

This script requires multiple core spatial packages, each of which is loaded below. Key variables and paths are also created.

```{r}
#| label: load packages

#use pacman function to load and install (if required) all other packages
pacman::p_load(tidyverse, glue, here, sf, tmap, janitor, dataaimsr, grid)

#load in the custom function used to create the read and write folders for the script
source(here("functions/script_setup.R"))

#run the function to create the folders and paths
script_setup()

#set project crs
#proj_crs <- params$project_crs

proj_crs <- "EPSG:7844"

#read in the custom function to clean column names into our specific style
source(here("functions/name_cleaning.R"))

```

# Load Data

Next we load in the data, refer to the README in the GitHub repo if data is missing.

```{r}
#| label: load in coral data

#get only the dry tropics area
dry_tropics <- n3_region |> 
  name_cleaning() |> 
  filter(Region == "Dry Tropics")

#get the marine zone for the dry tropics region
dt_marine <- n3_region |> 
  filter(Region == "Dry Tropics",
         Environment == "Marine", 
         BasinOrZone != "Offshore") |> 
  mutate(WatercourseOrGeographicArea = case_when(WatercourseOrGeographicArea == "Open Coastal" ~ "H.Open Coastal",
                                    WatercourseOrGeographicArea == "Enclosed Coastal" ~ "H.Enclosed Coastal",
                                    T ~ WatercourseOrGeographicArea))

#get specifically a land version for the background
dt_land <- dry_tropics |> 
         filter(Environment != "Marine") |> 
  group_by(BasinOrZone) |> 
  summarise(geom = st_union(geom)) |> 
  ungroup() |> st_cast() |> st_make_valid()

#read in reef data from the aims gis package and update crs
reefs <- get(data("gbr_feat", package = "gisaimsr")) |> 
  name_cleaning() |>
  filter(FeatName == "Reef") |> 
  st_transform(proj_crs)

#load in custom reef outlines to supplement the main data
custom_reefs <- st_read(glue("{data_path}/custom_reefs.gpkg")) |> 
  name_cleaning() |> 
  st_transform(proj_crs)

st_write(reefs, "reef_outlines.gpkg")

#create a point of interest for townsville from the poi dataset
tsv <- poi |> filter(PlaceName == "Townsville")

```

# Maipulate Data

The names in the reef dataset are too excessive and can visually clutter our map, in this code chunk we clean up the reef names.

```{r}
#| label: clean up data

#crop reefs to only in Dry Tropics region and take only polygons
temp_reefs <- st_intersection(reefs, st_union(dt_marine)) |> 
  st_collection_extract("POLYGON")

#create list for reefs (old and new names) and x and y mods, the order is: Old name = New name, Xmod, Ymod
reef_list<- list("Alma Bay" = c("Alma Bay", 1.8, 0),
                 "Magnetic Island Reef (No 1)" = c("Nelly Bay", 1.7, 0),
                 "Magnetic Island Reef (No 2)" = c("Geoffrey Bay (2 sites)", 3, -0.5),
                 "Magnetic Island Reef (No 3)" = c("Florence Bay", 2.4, 0),
                 "Middle Reef" = c("Middle Reef", 2.5, 0),
                 "Havannah Reef" = c("Havannah South", -3, 0),
                 "Havannah_N" = c("Havannah North", 3, 0),
                 "Palms West 1" = c("Palms West 1", -2.4, 0),
                 "Pelorus (North Palm or Yanooa) Reef" = c("Palms West 2", -2.4, 0.5),
                 "Orpheus (Goolboddi) Island Reef (No 5)" = c("Palms East", 2.2, 0),
                 "Pandora Reef" = c("Pandora North", 2.4, 0.2),
                 "Pandora Reef South" = c("Pandora South", -2.4, -0.2))

#filter by name to get reefs we want, add missing reefs and missing columns
dt_reefs <- temp_reefs |> filter(GbrName %in% names(reef_list)) |> select(GbrName) |>
  rbind(custom_reefs) |> 
  mutate(Reef = NA, Xmod = NA, Ymod = NA)

#loop over the list and add each item
for (i in 1:length(reef_list)){
  
  dt_reefs <- dt_reefs |> mutate(Reef = case_when(GbrName == names(reef_list)[i] ~ reef_list[[i]][1], T ~ Reef),
                                 Xmod = case_when(GbrName == names(reef_list)[i] ~ as.numeric(reef_list[[i]][2]), T ~ Xmod),
                                 Ymod = case_when(GbrName == names(reef_list)[i] ~ as.numeric(reef_list[[i]][3]), T ~ Ymod),
                                 Zone = case_when(GbrName %in% names(reef_list)[1:5] ~ "Cleveland Bay", T ~ "Halifax Bay"))

}

```

# Create Map

And finally we can produce the required maps to be used in our technical report style.

```{r}
#| label: create map

#load in custom function to create the water related layers of the map
source(here("functions/maps_water_layer.R"))

#load in custom function to create the inset map
source(here("functions/maps_inset_layer.R"))

#create a background of the burdekin region for the inset map
dt_background <- dry_tropics |> 
  group_by(BasinOrZone) |> 
  summarise(geom = st_union(geom))

#create target vector to loop lover
target_vect <- unique(dt_marine$BasinOrZone)

#create a similar vect for land basins to help with filtering
land_vect <- rev(unique(dt_land$BasinOrZone))

for (i in 1:length(target_vect)){
  
  #filter sites
  filtered_sites <- dt_reefs |> filter(Zone == target_vect[i])
  
  #create an invisible buffered version of the sites to help position the map
  buff_sites <- st_buffer(filtered_sites, as.numeric(sprintf("0.%s", i)))
  
  #define focus area for overall marine zone and rename column
  focus_marine <- dt_marine |> 
    filter(BasinOrZone == target_vect[i]) |> 
    rename("Geographic Area: All" = WatercourseOrGeographicArea)
  
  #create the base water layer for the specific basin
  maps_water_layer(stream_order = c(2,9), water_lines = T, water_polygons = T, 
                   basin_or_zone = land_vect[i], enviro = c("Freshwater", "Estuarine"), fast = T)
  
  #create the background layer using the broader dry tropics backdrop
  maps_inset_layer(supplied_sf_1 = focus_marine, supplied_sf_2 = filtered_sites, 
                   use_bbox_1 = FALSE, use_bbox_2 = TRUE,
                   background = dt_background, aspect = 1)

  #create the main map
  map <- tm_shape(qld) +
    tm_polygons(fill = "grey80") +
    tm_shape(buff_sites, is.main = T) +
    tm_polygons(fill_alpha = 0, 
                col_alpha = 0) +
    tm_shape(dt_marine) +
    tm_borders(col = "black") +
    tm_shape(dt_land) +
    tm_polygons(fill = "grey90", 
                col = "black") +
    tm_shape(focus_marine) +
    tm_polygons(fill = "Geographic Area: All", 
                fill_alpha = 0.5,
                fill.scale = tm_scale_categorical(values = "brewer.pastel1"),
                col = "black") +
    water_map +
    tm_shape(tsv) +
    tm_symbols(size = 0.3, 
               fill = "white", 
               col = "black", 
               lwd = 2, 
               shape = 23) +
    tm_text("PlaceName", 
            size = 0.8,
            options = opt_tm_text(shadow = T)) +
    tm_shape(filtered_sites) +
    tm_polygons(fill = "Reef", 
                fill.scale = tm_scale_categorical(values = "brewer.pastel2"),
                fill.legend = tm_legend(show = FALSE),
                col = "black") +
    tm_text("Reef", 
            options = opt_tm_text(shadow = T),
            xmod = "Xmod", 
            ymod = "Ymod", 
            size = 0.8) +
    tm_layout(legend.bg.color = "white", 
              legend.position = c("left", "bottom"),
              asp = 1)
  
  #and save
  tmap_save(map, glue("{output_path}/inshore-marine_{tolower(target_vect[i])}-coral.png"), 
            insets_tm = inset_map, insets_vp = inset_viewport)

}

```

Below is an interactive example of the maps created.

```{r}
#| label: interactive map example

tmap_mode("view")

map

```

# Session Info {#sec-sessioninfo}

Below is the session info at the time of rendering this script. Of greatest importance is to note the R version, and the "other attached packages" as these are the most significant drivers of success/failure. It is also good to check the "attached base packages" and "loaded via a namespace" packages as well. To check your session info use `sessionInfo()`.

```{r}
#| label: show session info

sessionInfo()

```
